---
title: "homework5"
author: "Andy Powers"
format: html
---

# Homework 5

## Context

### Datasets

Our dataset comes from the UCI ML repository and describes secondary education in two Portuguese schools.

[Data dictionary](https://archive.ics.uci.edu/ml/datasets/Student+Performance)

[Dataset](https://archive.ics.uci.edu/dataset/320/student+performance)

Dictionary text:

-   1 school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)

-   2 sex - student's sex (binary: 'F' - female or 'M' - male)

-   3 age - student's age (numeric: from 15 to 22)

-   4 address - student's home address type (binary: 'U' - urban or 'R' - rural)

-   5 famsize - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3)

-   6 Pstatus - parent's cohabitation status (binary: 'T' - living together or 'A' - apart)

-   7 Medu - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary - education or 4 – higher education)

-   8 Fedu - father's education (numeric: 0 - none, 1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary - education or 4 – higher education)

-   9 Mjob - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), - 'at_home' or 'other')

-   10 Fjob - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), - 'at_home' or 'other')

-   11 reason - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other')

-   12 guardian - student's guardian (nominal: 'mother', 'father' or 'other')

-   13 traveltime - home to school travel time (numeric: 1 - \<15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - \>1 hour)

-   14 studytime - weekly study time (numeric: 1 - \<2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - \>10 hours)

-   15 failures - number of past class failures (numeric: n if 1\<=n\<3, else 4)

-   16 schoolsup - extra educational support (binary: yes or no)

-   17 famsup - family educational support (binary: yes or no)

-   18 paid - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)

-   19 activities - extra-curricular activities (binary: yes or no)

-   20 nursery - attended nursery school (binary: yes or no)

-   21 higher - wants to take higher education (binary: yes or no)

-   22 internet - Internet access at home (binary: yes or no)

-   23 romantic - with a romantic relationship (binary: yes or no)

-   24 famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)

-   25 freetime - free time after school (numeric: from 1 - very low to 5 - very high)

-   26 goout - going out with friends (numeric: from 1 - very low to 5 - very high)

-   27 Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)

-   28 Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)

-   29 health - current health status (numeric: from 1 - very bad to 5 - very good)

-   30 absences - number of school absences (numeric: from 0 to 93)

These grades are related with the course subject, Math or Portuguese:

-   31 G1 - first period grade (numeric: from 0 to 20)

-   31 G2 - second period grade (numeric: from 0 to 20)

-   32 G3 - final grade (numeric: from 0 to 20, output target)

### Libraries

```{r}
library(tidyverse)
library(psych)
```

## Read and Modify Data

### Read and Join

First, reading in using the provided R code:

```{r}
d1 <- read.table("student-mat.csv",sep=";",header=TRUE)
d2 <- read.table("student-por.csv",sep=";",header=TRUE)

d3 <- merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
print(nrow(d3)) # 382 students

```

Now, reading in using `tidyverse` methods.

```{r}
mat_data <- read_csv2(file = "student-mat.csv")
por_data <- read_csv2(file = "student-por.csv")
all_data <- inner_join(mat_data,por_data,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))
```

There is a warning announced, referencing a many-to-many relationship between x and y. Trying again, joining on all variables except `G1`,`G2`,`G3`,`paid`,`absences`. Will use this result for all further exercises. Rather than listing all 28 desired vars for join, manually, I will first join on a selection of the two tables without the noted five fields. Then, I will join this result with the original mat and por data to get the additional fields, adjusting the names to include the course type as a suffix (G1 -> G1.mat, G1.por, etc.).

```{r}
#join the subset first
all_data <- inner_join(
  select(mat_data,-c(G1,G2,G3,paid,absences)),
  select(por_data,-c(G1,G2,G3,paid,absences))
  )

#add in the missing fields from mat and rename
all_data <- 
  left_join(
    all_data,
    mat_data
    ) |>
  rename_with(~ paste(.x,".mat",sep=""),
              .cols = c(G1,G2,G3,paid,absences)
  )

#similar activity for fields from por
all_data <- 
  left_join(
    all_data,
    por_data
    ) |>
  rename_with(~ paste(.x,".por",sep=""),
              .cols = c(G1,G2,G3,paid,absences)
  )

```

### Factoring Vars of Interest

I'd like to study the relationships of grades with these categorical variables:

1.  Pstatus: parent's cohabitation status (binary: 'T' - living together or 'A' - apart)
2.  freetime: free time after school (numeric: from 1 - very low to 5 - very high)
3.  famrel: quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
4.  romantic: with a romantic relationship (binary: yes or no)

*Important note: For some of these fields, I am changing values from 1-5 to labels, but I do not have sufficient access to the original terminology used by survey participants for each value 1,2,3,4,5. I have the label for the 1 and the 5, only. I am adding a typical label for the middle responses,* **but these are a guess on my part and not suitable for strict interpretation as listed***.*

So, we will convert these to `factor` fields, for all three tibbles (math, portuguese, combined). To make this modular and clean, I will define a function to do this, then call it for each tibble.

```{r}
#writing a helper function for this purpose

#factorize_4_fields(myTibble)
#Takes a tibble as input, modifies the four fields noted above into factors, returns the modified tibble.
#Expects only a tibble of the por/mat_data dataset, no error checking.

factorize_4_fields <- function(myTibble) {
  myTibble <- 
    myTibble |>
    mutate(
      Pstatus = factor(Pstatus,
                       levels=c("A","T"),
                       labels=c("apart","together")
                       ),
      freetime = factor(freetime,
                        levels=c(1,2,3,4,5),
                        labels=c("1-very low", "2-low", "3-average", "4-high", "5-very high")
                        ),
      famrel = factor(famrel,
                      levels=c(1,2,3,4,5),
                      labels=c("1-very bad", "2-bad", "3-average", "4-good", "5-excellent")
                      ),
      romantic = factor(romantic,
                        levels=c("yes","no"),
                        labels=c("romance","no romance")
                        )
    )
  return(myTibble)
}

```

Now, running this factor mutation for each tibble:

```{r}

all_data <- 
  all_data |>
  factorize_4_fields()

mat_data <- 
  mat_data |>
  factorize_4_fields()

por_data <- 
  por_data |>
  factorize_4_fields()

```

## Summarize Data with Basic EDA

### Data Review

First, we will look at the data and evaluate whether it makes basic sense. We will also look for missing or malformed values, etc.

#### Structures, types, values

```{r}
str(all_data)
```

Looking at results of a structure `str()` function, the field names and their values look reasonable. They match the descriptions of the data dictionary. However, we need to adjust the numeric and categorical field types, as many of these fields have a numeric type currently but are truly categorical in nature. To be efficient, I will try to adjust *all* numeric fields *except* those I specify, to be coerced to character strings.

```{r}
all_data <- 
  all_data |>
  mutate(across(
           where(is.numeric)
           & !ends_with(".mat")
           & !ends_with(".por")
           & -c(failures,age),
           as.character #transform from numeric to char
           )
         )
```

#### Prior data transforms

I am concerned that there are only 320 records in the combined data, while both math and portuguese datasets had record volumes exceeding this amount (395, 649, respectively). Thinking through our join methodology, since we used an inner join, this gives us only the students who had data in both courses. That's not a bad approach and is even necessary for some analyses, so this is acceptable.

#### Numeric summary

I'll run psych::describe() to get a sense of the values and their centers, spread, to ensure that looks reasonable. Only the numeric ones.

```{r}
describe(select(all_data,is.numeric))
```

These all look reasonable for their representations of failed classes, number of absences, and grades.

#### Missing values

Let's check for any NA, though I see the original datasets used here were very clean.

```{r}
colSums(is.na(all_data))
```

All clean, no missing values in any field.

### EDA for Categorical Vars

#### 1,2,3-Way Contingency Tables

These are easy to create using `table()` so there isn't much preamble needed. Interpretation afterward. These are built with BaseR.

```{r}
table(all_data['Pstatus'])
```
This one-way contingency table simply tells us that most of the studied students' parents lived together (289) while a smaller count did not (31). The numbers are basic counts of each element of the field Pstatus.

```{r}
table(all_data[c('Pstatus','romantic')])
```
This two-way table shows the split of the prior table by the romance / no romance values. Of the 31 students whose parents lived separately, 13 students were in a romantic relationship. That's interesting, as a percentage of the total, because the share of romance / no romance is much closer for the *apart* group than the share for the *together* group.

```{r}
table(all_data[c('Pstatus','romantic','famrel')])

```

The three-way contingency table with a 5-element field is much harder to interpret. Looking at the *1-very bad* relationship group, 2 students have separated parents and are in a romantic relationship. While the volumes for the *1-very bad* group are low, it is the only level of family relationship where the total romance-active students exceeds that of no-romance students.

#### Conditional 2-Way Table

First, creating this via `table()` and a filter.

```{r}
table_cond_2way_a <- filter(all_data,famrel=="1-very bad")
table(table_cond_2way_a[c('Pstatus','romantic')])
```
This looks like our sample above, reflecting only the counts of the filtered student data.

Creating this another way, I'll make the 3-way table and then subset it:

```{r}
table_cond_2way <- table(all_data[c('Pstatus','romantic','famrel')])
table_cond_2way[,,1]
```
Looks good.

#### 2-Way Contingency Table with dplyr

I will make these using `group_by()` and `summarize()` to test other methods.

```{r}
table_dplyr <- 
  all_data |>
  group_by(Pstatus,romantic) |>
  summarize(count = n())
table_dplyr
```

The values match but the display format does not. `pivot_wider` will adjust this to mimic the prior result from BaseR, the typical display for a contingency table.

```{r}
table_dplyr <- pivot_wider(table_dplyr,names_from = "romantic",values_from = "count")
table_dplyr
```

Matches the original display of `table()`.

#### Bar Graphs

Now, using `ggplot2` I will display a stacked bar graph and side-by-side variant of the same. Good labels for the axes and a title, too.

```{r}
g <- ggplot(all_data,
            aes(x=Pstatus,
                fill=romantic
                )
            )
g + 
  geom_bar(position="stack") +
  labs(
    x="Cohabitation Status of Parents",
    y="Count of Students",
    fill="Student Romance",
    title = "Cohabitation State of Parents & Student Romance Status")
```

```{r}
g + 
  geom_bar(position="dodge") +
  labs(
    x="Cohabitation Status of Parents",
    y="Count of Students",
    fill="Student Romance",
    title = "Cohabitation State of Parents & Student Romance Status") 

```
### EDA for Numeric Vars and Groups

#### Center & Spread for 3 Vars

First, calculating basic measures of center and spread for three vars:

```{r}
all_data |>
  select(age,G3.mat,G3.por) |>
  summarize(across(everything(),
                   list("mean"=mean,
                        "median"=median,
                        "sd"=sd,
                        "IQR"=IQR),
                   .names="{.col}-{.fn}"
                   )
            )
```
Looking at `age`, we know most students are about 16, +/- 1. Now, let's see the summary results for these students when we only assess those with a romantic relationship.

```{r}
all_data |>
  filter(romantic=="romance") |>
  select(age,G3.mat,G3.por) |>
  summarize(across(everything(),
                   list("mean"=mean,
                        "median"=median,
                        "sd"=sd,
                        "IQR"=IQR),
                   .names="{.col}-{.fn}"
                   )
            )
```
This result shows higher typical ages for students in a romantic relationship, along with noticeably lower final math grades but comparable language grades.

#### By a Single Grouping Variable

Now, let's compare these results for both of the romance states of students.

```{r}
all_data |>
  group_by(romantic) |>
  select(age,G3.mat,absences.mat) |>
  summarize(across(everything(),
                   list("mean"=mean,
                        "median"=median,
                        "sd"=sd,
                        "IQR"=IQR),
                   .names="{.col}-{.fn}"
                   )
            )
```

There are higher numbers of absences associated with those students in romantic relationships, though with a much larger spread (and a smaller sample size).

#### By 2 Grouping Variables

Now, let's group additionally by `famrel`.

```{r}
all_data |>
  group_by(romantic,famrel) |>
  select(age,G3.mat,G3.por) |>
  summarize(across(everything(),
                   list("mean"=mean,
                        "median"=median,
                        "sd"=sd,
                        "IQR"=IQR),
                   .names="{.col}-{.fn}"
                   )
            )
```

Well, with one of my variables having 5 levels, this output is hard to read. Further, I wish strongly that I could add in the count of students in each grouping but I can't figure out how. With the levels of `famrel` ordered as factors, we can see a lightly-apparent trend that `age` is positively correlated to quality of family relationships.

#### Correlation Matrix for All Numerics

This may be hard to read, but now we will make a correlation matrix for all numeric variables. I don't recall this being in the lesson, but I reviewed how to use cor() to do this.

```{r}
cor(all_data[sapply(all_data,is.numeric)]) 
```

#### Histogram, Kernel Density, Boxplot

I will create these for the numeric variables `absences.mat` and `G3.mat`, for categorical variable `freetime`.

```{r}
#c + geom_histogram(binwidth = 5)
#x, y, alpha, color, fill, linetype, size, weight 
g <- ggplot(all_data,
            aes(x=Pstatus,
                fill=romantic
                )
            )
g + 
  geom_bar(position="stack") +
  labs(
    x="Cohabitation Status of Parents",
    y="Count of Students",
    fill="Student Romance",
    title = "Cohabitation State of Parents & Student Romance Status")

g2 <- ggplot(all_data,
              aes(x=absences.mat,
                  fill=G3.mat)
              )

g2 + 
  geom_histogram(binwidth=3)
```

```{r}
#c + geom_density(kernel = "gaussian")
g3 <- ggplot(all_data,
              aes(x=absences.mat)
)

g3 + 
  geom_density(
    kernel="gaussian",
    alpha=0.5,
    aes(
      fill=freetime
      )
    ) +
  labs
```

